Sablefish equations

### 2024-08-22 ms/equations.tex
tables with concise dims

\section{Movement model: decision point}

@@

\noindent Upshot: I recommend the compact notation because it is concise. I believe it gives a better `sense' of the model on a quick scan, and a more direct route to understanding the nuances on a deep dive.

@@

\section{Movement model: compact indexes}

@@

\subsection{Tables}

% Model indexes
\begin{table}[ht]
  \centering
  \caption{Movement model indexes}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-indexes}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Condition} & \textbf{Dimension} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    % Index limits
    $S$ & $= 3$ or $6$ & Scalar & Number of spatial regions & Index limit \\
    $T$ & $\in \mathbb{Z}^{+}$ & Scalar & Number of model time steps & Index limit \\
    $K$ & $= 6$ & Scalar & Number of time steps per year & Index limit \\
    $L$ & $= 1$ or $2$ & Scalar & Number of release size classes & Index limit \\
    \midrule
    % Index variables
    $\boldsymbol{s}$ & $= (s_0, s)$ & Vector & Spatial index vector & Vector index \\
    $\boldsymbol{t}$ & $= (t_0, t)$ & Vector & Temporal index vector & Vector index \\
    $\boldsymbol{s}_0$ & $= (s_0, s_0)$ & Vector & Initial spatial index vector & Vector index \\
    $\boldsymbol{t}_0$ & $= (t_0, t_0)$ & Vector & Initial temporal index vector & Vector index \\
    \midrule
    $s$ & $\in \left[1 \, .. \, S \right]$ & Scalar & Spatial region & Index \\
    $t$ & $\in \left[1 \, .. \, T \right]$ & Scalar & Time step & Index \\
    $s_0$ & $\in \left[1 \, .. \, S \right]$ & Scalar & Initial spatial region & Index \\
    $t_0$ & $\in \left[1 \, .. \, T \! - \! 1 \right]$ & Scalar & Initial model time step & Index \\
    $k$ & $\in \left[1 \, .. \, K \right]$ & Scalar & Time step within year & Index \\
    $l$ & $\in \left[1 \, .. \, L \right]$ & Scalar & Release size class & Index \\
    \bottomrule
  \end{tabular}
\end{table}

\noindent Note: needs updates to language, possibly symbols (C to Q; s to lambda; theta to W?)

We modeled sablefish movement as a hidden Markov process \cite[][]{langrock-2012-flexible-practical} observed by tag recoveries. Sablefish entered the model as counts of tag releases identified by release region, time-step, and size class. The time-step in our analysis was one quarter-year. For a given release time-step and size class, the initial abundances of tagged sablefish were represented by a square diagonal matrix

% Survival
\begin{equation}
  \label{eq:abundance-initial}
  \mathrm{diag} \! \left[\boldsymbol{A}_{i,d=1,l}\right] = \boldsymbol{R}_{i,l} \left(1 - \nu \right)
\end{equation}

where row corresponded to release region and diagonal elements were counts of tags released in each region diminished by the initial tag loss rate $\upsilon$. Subsequent tag abundances from the same release time-step and size class were defined recursively as

% Abundance
\begin{equation}
    \label{eq:abundance}
    \boldsymbol{A}_{i,d,l} = \boldsymbol{A}_{i,d-1,l} \, \boldsymbol{\Gamma}_{n[i,d-1],l}
\end{equation}

\noindent where $n$ was some time-step after release, $\boldsymbol{\Gamma}_n$ acted as a transition matrix, and $\boldsymbol{A}_n$ encoded the projected tag abundances at the beginning of time-step $n$ with rows indicating the release region and columns indicating the destination region.

What we refer to as the transition matrix $\boldsymbol{\Gamma}_n$ combined survival and movement rates as the matrix product

% Transition
\begin{equation}
    \label{eq:transition}
    \boldsymbol{\Gamma}_{n,l} = \boldsymbol{S}_{n,l} \, \boldsymbol{\Delta}_{l}
\end{equation}

\noindent where $\boldsymbol{S}_t$ was a diagonal matrix of step-wise survival rates in year $t$, $\boldsymbol{\Delta}_t$ was a square matrix of step-wise movement rates in year $t$, and $t[n]$ indexed year as a function of model step. While $\boldsymbol{\Delta}_t$ was a proper right stochastic matrix in the sense that its elements were probabilities and rows summed to one, $\boldsymbol{\Gamma}_n$ was not, because mortality removed some tagged sablefish from the model during each time-step. Consequently, mortality (including tag loss) formed an absorbing Markov state that could not be departed, in addition to the eight discrete non-absorbing states defined by the management regions. We chose to encode the mortality state implicitly via removal for clarity rather than assign an additional row and column to $\boldsymbol{\Gamma}_n$, and we refer to $\boldsymbol{\Gamma}_n$ as a transition matrix for convenience with the understanding that its rows did not sum to one.

Sablefish survival rates depended on mortality and tag loss as

% Survival
\begin{equation}
  \label{eq:survival}
  \mathrm{diag} \! \left[\boldsymbol{S}_{n,l}\right] = 
    \exp\!{\left[-\boldsymbol{s}_l \boldsymbol{f}_n - \boldsymbol{m} - \eta \right]}
\end{equation}

\noindent where $\boldsymbol{f}_n$ and $\boldsymbol{m}$ were row vectors of step-wise fishing and natural mortality, and $\eta$ was step-wise ongoing tag loss including tag-induced mortality.

We modeled expected tag recoveries by the matrix product of tagged sablefish abundance and a step-wise observation rate. For a given release time-step and size class this was defined as

% Expected
\begin{equation}
  \label{eq:expected}
  \boldsymbol{\widehat{C}}_{i,d,l} = \boldsymbol{A}_{i,d,l} \boldsymbol{B}_{n[i,d],l}
\end{equation}

\noindent where $\boldsymbol{\widehat{C}}_{i,d,l}$ was a square matrix of expected tag recoveries with rows indicating release region and columns indicating recovery region, and $\boldsymbol{B}_{n,l}$ was a diagonal matrix of observation rates defined by the element-wise vector product

% Observation
\begin{equation}
  \label{eq:observation}
  \mathrm{diag} \! \left[\boldsymbol{B}_{n,l}\right] = 
    \boldsymbol{\theta} \left(1 - \exp\!{\left[-\boldsymbol{s}_l \boldsymbol{f}_{n} \right]} \right) 
\end{equation}

\noindent where $\boldsymbol{\theta}$ was a row vector of per-tag reporting rates and $\boldsymbol{f}_n$ was a vector of step-wise fishing mortality rates.

For a given release size class, we defined annual sablefish movement rates by a matrix power of the corresponding step-wise movement rates

% Movement rate
\begin{equation}
    \label{eq:movement}
    \boldsymbol{P}_{l} = \boldsymbol{\Delta}_{l}^K
\end{equation}

\noindent where $\boldsymbol{P}_{l}$ was the square matrix of annual movement rates, and $K$ was the number of time-steps per year.

% The matrix of step-wise movement rate deviations $\boldsymbol{\Lambda}_t = \boldsymbol{\Delta}_t - \boldsymbol{\bar{\Delta}}$ from the step-wise mean movement rates $\boldsymbol{\bar{\Delta}}$ was constrained by an autoregressive process on its diagonal

% Timevary
%\begin{equation}
%  \label{eq:timevary}
%  \mathrm{diag}\!\left[\boldsymbol{\Lambda}_t\right] = \mathrm{diag}\!\left[\boldsymbol{\Lambda}_{t-1}\right]\! \boldsymbol{\Psi} + \boldsymbol{\epsilon}_t, \, \boldsymbol{\epsilon}_t \sim \mathrm{N}\!\left[\boldsymbol{0},\,\boldsymbol{\Sigma}_\Lambda\right]
%\end{equation}

%\noindent where $\boldsymbol{\Psi}$ was a diagonal matrix of autocorrelation coefficients, and $\boldsymbol{\Sigma}_\Lambda$ was a diagonal variance matrix.

We used the NB2 parameterization of the negative binomial distribution with probability mass function

% Sampling
\begin{equation}
  \label{eq:sampling}
  \mathrm{NB2} \! \left[ C_* \mid \widehat{C}_*,\, \phi \right] = \binom{C_* + \phi - 1}{C_*} \left(\frac{\widehat{C}_*}{\widehat{C}_* + \phi}\right)^{C_*} \left( \frac{\phi}{\widehat{C}_* + \phi} \right)^{\phi}
\end{equation}

\noindent where $C_*$ and $\widehat{C}_*$ were the observed and expected tag recovery counts corresponding to a given release region, time-step and length class, and recovery region and time-step.

% Model indexes
\begin{table}[ht]
  \centering
  \caption{Movement model indexes}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-indexes-0}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Condition} & \textbf{Dimension} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    % Index limits
    $X$ & $\in \mathbb{Z}^{+}$ & Scalar & Number of geographic regions & Index limit \\
    $T$ & $\in \mathbb{Z}^{+}$ & Scalar & Number of years & Index limit \\
    $K$ & $\in \mathbb{Z}^{+}$ & Scalar & Number of time steps per year & Index limit \\
    $N$ & $ = TK $ & Scalar & Number of time steps & Index limit \\
    $D$ & $\in \mathbb{Z}^{+}$ & Scalar & Maximum time steps at liberty & Index limit \\
    $L$ & $\in \mathbb{Z}^{+}$ & Scalar & Number of release size classes & Index limit \\
    \midrule
    % Index variables
    $x, \, y$ & $\in \left[1 \, .. \, X \right]$ & Scalar & Geographic region & Index \\
    $t$ & $\in \left[1 \, .. \, T \right]$ & Scalar & Year & Index \\
    $n$ & $\in \left[1 \, .. \, N \right]$ & Scalar & Time step & Index \\
    $i$ & $\in \left[1 \, .. \, N\!-\!1 \right]$ & Scalar & Release time step & Index \\
    $k$ & $\in \left[1 \, .. \, K \right]$ & Scalar & Time step within year & Index \\
    $d$ & $\in \left[1 \, .. \, D \right]$ & Scalar & Time step at liberty & Index \\
    $l$ & $\in \left[1 \, .. \, S \right]$ & Scalar & Release size class & Index \\
    \bottomrule
  \end{tabular}
\end{table}

% Data
\begin{table}[ht]
  \centering
  \caption{Data}
  \renewcommand\arraystretch{1.2}
  \label{tab:data}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Condition} & \textbf{Dimension} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    $\bm{R}$ & $\in \mathbb{Z}^{+,0}$ & $[N\!-\!1, L][X]$ & Sablefish tags released (counts) & Stepwise \\
    $\bm{C}$ & $\in \mathbb{Z}^{+,0}$ & $[N\!-\!1, D, L][X, X]$ & Sablefish tags recovered (counts) & Stepwise \\
    \bottomrule
  \end{tabular}
\end{table}

% Model derived quantities
\begin{table}[ht]
  \centering
  \caption{Movement model derived quantities}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-derived}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Condition} & \textbf{Dimension} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    $\bm{A}$ & $\in \mathbb{R}^{+,0}$ & $[N\!-\!1, D, L][X, X]$ & Sablefish tag abundance (numbers) & Stepwise \\
    $\boldsymbol{\widehat{C}}$ & $\in \mathbb{R}^{+,0}$ & $[N\!-\!1, D, L][X, X]$ & Sablefish expected tags recovered (numbers) & Stepwise \\
    $\boldsymbol{\Gamma}$ & $\in \left[0, 1 \right]$ & $[N, L][X, X]$ & Sablefish transition rates & Stepwise \\
    $\bm{S}$ & $\in \left[0, 1 \right]$ & $[N, L][X, X]$ & Sablefish survival rates & Stepwise \\
    $\bm{B}$ & $\in \left[0, 1 \right]$ & $[N, L][X]$ & Sablefish tag observation rates & Stepwise \\
    \midrule
    $\bm{P}$ & $\in \left[0, 1 \right]$ & $[L][X, X]$ & Sablefish movement rates & Annual \\
    $\bm{F}$ & $\in \mathbb{R}^{+,0}$ & $[T, X]$ & Sablefish instantaneous fishing mortality rates & Annual \\
    $\bm{M}$ & $\in \mathbb{R}^{+,0}$ & $[T, X]$ & Sablefish instantaneous natural mortality rates & Annual \\
%    \midrule
%    $\boldsymbol{\Lambda}$ & $\in \left[0, 1 \right]$ & $[L][X, X]$ & Sablefish movement rate deviations & Stepwise \\    
    \bottomrule
  \end{tabular}
\end{table}

% Model parameters
\begin{table}[ht]
  \centering
  \caption{Movement model parameters}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-parameters-0}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Condition} & \textbf{Dimension} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    $\boldsymbol{\Delta}$ & $\in \left[0, 1 \right]$ & $[L][X, X]$ & Sablefish movement rates & Stepwise \\
    $\bm{f}$ & $\in \mathbb{R}^{+,0}$ & $[N, X]$ & Sablefish instantaneous fishing mortality rates & Stepwise \\
    $\bm{s}$ & $\in \left[0, 1 \right]$ & $[L, X]$ & Sablefish fisheries size selectivity & Per fish \\
    $\bm{m}$ & $\in \mathbb{R}^{+,0}$ & $[X]$ & Sablefish instantaneous natural mortality rates & Stepwise \\
    $\boldsymbol{\theta}$ & $\in \left[0, 1 \right]$ & $[X]$ & Sablefish tag reporting rates & Per tag \\
    $\eta$ & $\in \mathbb{R}^{+,0}$ & Scalar & Sablefish instantaneous ongoing tag loss rate & Stepwise \\
    $\nu$ & $\in \left[0, 1 \right]$ & Scalar & Sablefish initial tag loss rate & Per tag \\    
    $\phi$ & $\in \mathbb{R}^{+}$ & Scalar & Negative binomial dispersion & NB2 \\
%    \midrule
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\section{Notation: decision point}

@@

\noindent Upshot: I recommend the compact notation because it is concise. I believe it gives a better sense of the model on a quick scan, and a more direct route to understanding the nuances on a deep dive.

@@

@@

\newpage
\section{Tables}

% Model indexes
\begin{table}[ht]
  \centering
  \caption{Model indexes}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-indexes}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Value} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    % Index limits
    $S$ & $= 3$ or $6$ & Number of spatial regions & Limit \\
    $T$ & $\in \mathbb{Z}^{+}$ & Number of model time steps & Limit \\
    $K$ & $= 6$ & Number of time steps per year & Limit \\
    $L$ & $= 1$ or $2$ & Number of release size classes & Limit \\
    \midrule
    % Index variables
    $\boldsymbol{s}$ & $= (s_0, s)$ & Spatial index vector & Index \\
    $\boldsymbol{t}$ & $= (t_0, t)$ & Temporal index vector & Index \\
    $\boldsymbol{s}_0$ & $= (s_0, s_0)$ & Initial spatial index vector & Index \\
    $\boldsymbol{t}_0$ & $= (t_0, t_0)$ & Initial temporal index vector & Index \\
    \midrule
    $s$ & $\in \left[1 \, .. \, S \right]$ & Spatial region & Index \\
    $t$ & $\in \left[1 \, .. \, T \right]$ & Time step & Index \\
    $s_0$ & $\in \left[1 \, .. \, S \right]$ & Initial spatial region & Index \\
    $t_0$ & $\in \left[1 \, .. \, T \! - \! 1 \right]$ & Initial model time step & Index \\
    $k$ & $\in \left[1 \, .. \, K \right]$ & Time step within year & Index \\
    $l$ & $\in \left[1 \, .. \, L \right]$ & Release size class & Index \\
    \bottomrule
  \end{tabular}
\end{table}

% Data
\begin{table}[ht]
  \centering
  \caption{Data}
  \renewcommand\arraystretch{1.2}
  \label{tab:data}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Value} & \textbf{Dim} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    $\boldsymbol{y}_{\boldsymbol{t},l}$ & $\in \mathbb{Z}^{+,0}$ & $[S,S]$ & Sablefish tags recovered & Count \\
    $\boldsymbol{x}_{t_0,l}$ & $\in \mathbb{Z}^{+,0}$ & $[S,S]$ & Sablefish tags released & Count \\
    $\boldsymbol{F}^{\mathrm{assess}}_{\mathrm{year}}$ & $\in \mathbb{R}^{+,0}$ & $[S,S]$ & Annual sablefish fishing mortality rate & Inst. \\    
    \bottomrule
  \end{tabular}
\end{table}

% Model derived quantities
\begin{table}[ht]
  \centering
  \caption{Derived quantities}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-derived}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Value} & \textbf{Dim} & \textbf{Definition}  & \textbf{Type} \\
    \toprule
    $\boldsymbol{N}_{\boldsymbol{t},l}$ & $\in \mathbb{R}^{+,0}$ & $[S,S]$ & Sablefish tag abundance & Numbers \\
    $\boldsymbol{\mu}_{\boldsymbol{t},l}$ & $\in \mathbb{R}^{+,0}$ & $[S,S]$ & Expected tags recovered & Numbers \\
    $\boldsymbol{\Psi}_{t,l}$ & $\in \left[0, 1 \right]$ & $[S,S]$ & Sablefish tag recovery rate & Stepwise \\
    $\boldsymbol{\Lambda}_{t,l}$ & $\in \left[0, 1 \right]$ & $[S,S]$ & Sablefish tag survival rate & Stepwise \\    
    $\boldsymbol{P}_{\mathrm{block},l}$ & $\in \left[0, 1 \right]$ & $[S,S]$ & Sablefish movement rate & Yearly \\
    \bottomrule
  \end{tabular}
\end{table}

% Model parameters
\begin{table}[ht]
  \centering
  \caption{Model parameters}
  \renewcommand\arraystretch{1.2}
  \label{tab:model-parameters}
  \begin{tabular}{l l l l r}
    \toprule
    \textbf{Symbol} & \textbf{Value} & \textbf{Dim} & \textbf{Definition} & \textbf{Type} \\
    \toprule
    $\boldsymbol{\Gamma}_{\mathrm{block},l}$ & $\in \left[0, 1 \right]$ & $[S,S]$ & Sablefish movement rate & Stepwise \\
    $\boldsymbol{\lambda}_{l}$ & $\in \left[0, 1 \right]$ & $[S,S]$ & Sablefish fisheries size selectivity & Per fish \\    
    $\boldsymbol{\omega}_{k}$ &  $\in \left[0, 1 \right]$ & $[S,S]$ & Seasonal fishing mortality weight & Stepwise \\
    $\boldsymbol{F}_{\mathrm{year}}$ & $\in \mathbb{R}^{+,0}$ & $[S,S]$ & Annual sablefish fishing mortality rate & Inst. \\
    $\boldsymbol{M}$ & $\in \mathbb{R}^{+,0}$ & $[S,S]$ & Annual sablefish natural mortality rate & Inst. \\
    $\boldsymbol{W}$ & $\in \left[0, 1 \right]$ & $[S,S]$ & Sablefish tag reporting rate & Per tag \\        
    $\eta$ & $\in \mathbb{R}^{+,0}$ & Scalar & Annual sablefish tag loss rate & Inst. \\    
    $\nu$ & $\in \left[0, 1 \right]$ & Scalar & Sablefish initial tag loss rate & Per tag \\
    $\phi$ & $\in \mathbb{R}^{+}$ & Scalar & Negative binomial dispersion & NB2 \\
    \bottomrule
  \end{tabular}
\end{table}
